\vspace{1.0cm} \Large \bf Integrating influenza antigenic dynamics with molecular evolution

### 2013-07-24 response.tex

The authors need to define their nomenclature regarding clade / lineage / type / strain. For instance, they say: ``We find that 1 unit of antigenic drift within a clade increases incidence by 1.26 standard deviations, while 1 unit of drift in a clade of the same type decreases incidence by 0.61 standard deviations and 1 unit of drift in clades of differing type decreases incidence by 0.32 standard deviations. The term for across lineage Bt is small in magnitude.'' I had to read this many times before concluding that type refers to A or B, clade to refers to any of the four classifications. I am still not sure if lineage has a specific meaning or is simply a redundant word for clade.

The comparison in Table 1 is crucial, but the model descriptions are very hard to understand until after reading the Methods carefully. The difficulty is because the models are described in technical terms (dimensionality, location prior, serum effects, virus effects). The most confusing part is the location prior. My understanding (and if this is wrong then things really need to be clarified) is that uninformed refers to making the predictions just from HI data, drift refers to making the predictions from HI data + date of isolation, and diffusion / drift refers to making the predictions from HI data + date of isolation + phylogenetic relationship. If the models were named in terms of the data used to make the predictions rather than in terms of the prior, it would be easier to understand intuitively. For instance, a reader could then simply see, ``OK, they included the isolation date in the model and the predictions improved by this much.'' I recognize that the prior plays a role in addition to the data, but I feel that a more intuitive naming scheme should be considered with detailed descriptions in the Methods. There also might be a more intuitive way to denote serum effects and virus effects, although this is less confusing than the location prior.

\textbf{This is a very helpful suggestion.  We've included a new column in Table 1 labeled `Data' that specifies the combination of data sources that go into the model (HI vs HI/year vs HI/year/seq).  We've also clarified that serum and virus effects act to give a baseline expectation for titer when virus and serum have identical antigenic locations.}

### 2013-07-24 flux.tex
Fix #6 - model description.

We begin with Bayesian analogs of the models used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as $N$-dimensional locations.
In this case, `serum effects' are fixed to the maximum titers exhibited by particular ferret sera, and virus and serum locations follow an uninformative diffuse normal prior as described in the `Antigenic cartography' section of Materials and methods.
After comparing models of differing dimensions, Smith et al.\ \cite{Smith04} arrive at a 2D model as the preferred model for their data.
Smith et al.\ \cite{Smith04} implement a form of MDS, seeking to optimize virus and serum locations such that the sum of squared errors between expected and observed titers is minimized (eq.\ \ref{mds} of Materials and methods).
Here, in implementing BMDS, we provide a likelihood function for the probability of observing HI data given virus and serum locations (eq.\ \ref{bmds} of Materials and methods) and seek to estimate model parameters through Bayesian inference using Markov chain Monte Carlo (MCMC).
However, the basic antigenic model describing drop in HI titer as proportional to Euclidean distance between virus and serum locations is identical between these methods.

@@

%%% errortable %%%
\begin{table}[h]
	\centering
	\caption{\textbf{Average absolute prediction error of log$_2$ HI titer for test data across models and datasets.}}
	\label{errortable}	
	\makebox[\textwidth][c]{
	\begin{tabular}{ c c c c c c c c c } 
	\hline
	\multicolumn{5}{c}{} 	&	\multicolumn{4}{c}{Test error}  \\
	Model	&	Dimen 	& Location prior	& 	Serum effects 	&	Virus effects	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
	\hline	
	1		&	1D 		& Uninformed		&	Fixed 			&	None			&	1.35	&	0.94 	&	0.90	&	1.08	\\
	2		&	2D 		& Uninformed		&	Fixed 			&	None			&	0.91 	&	0.78 	&	0.82	&	0.90	\\
	3		&	3D 		& Uninformed		&	Fixed 			&	None			&	0.93 	&	0.80 	&	0.85	&	0.92	\\
	4		&	4D 		& Uninformed		&	Fixed 			&	None			&	0.98 	&	0.84 	&	0.90	&	0.97	\\	
	5		&	5D 		& Uninformed		&	Fixed 			&	None			&	1.04 	&	0.89 	&	0.98	&	1.04	\\	
	6		&	2D 		& Drift				&	Fixed 			&	None			&	0.91	&	0.75	&	0.77	&	0.83	\\
	7		&	2D 		& Diffusion/Drift	&	Fixed 			&	None			&	0.89	&	0.74	&	0.74	&	0.83	\\
	8		&	2D 		& Diffusion/Drift	&	Estimated 		&	None			&	0.77	&	0.73	&	0.66	&	0.75	\\	
	9		&	2D 		& Diffusion/Drift	&	Fixed 			&	Estimated		&	0.80	&	0.72	&	0.69	&	0.75	\\		
	10		&	2D 		& Diffusion/Drift	&	Estimated		&	Estimated		&	0.76	&	0.71 	&	0.64	&	0.72	\\
	\hline
	\end{tabular}	
	}
\end{table}

@@

We further extend the model by estimating the strength of overall reactivity of each serum rather than fixing this at its maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
We refer to these estimates as serum effects and virus effects, respectively.
We find that including these effects decreases test error further, resulting in comparable improvements from estimating virus and serum effects individually and giving an improvement of between 0.03 to 0.13 log$_2$ HI titers when estimating both virus and serum effects together (models 8--10; Table \ref{errortable}).

>>>>

We begin with Bayesian analogs of the models used by Smith et al.\ \cite{Smith04}, in which viruses and sera are represented as $N$-dimensional locations as described in the `Antigenic cartography' section of Materials and methods.
In this case, `serum effects' are fixed to the maximum titers exhibited by particular ferret sera and give the baseline expectation for titer when virus and serum are antigenically identical.
Virus and serum locations follow an uninformative diffuse normal prior. 
After comparing models of differing dimensions, Smith et al.\ \cite{Smith04} arrive at a 2D model as the preferred model for their data.
Smith et al.\ \cite{Smith04} implement a form of MDS, seeking to optimize virus and serum locations such that the sum of squared errors between expected and observed titers is minimized (eq.\ \ref{mds} of Materials and methods).
Here, in implementing BMDS, we provide a likelihood function for the probability of observing HI data given virus and serum locations (eq.\ \ref{bmds} of Materials and methods) and seek to estimate model parameters through Bayesian inference using Markov chain Monte Carlo (MCMC).
However, the basic antigenic model describing drop in HI titer as proportional to Euclidean distance between virus and serum locations is identical between these methods.

@@

%%% errortable %%%
\begin{table}[h]
	\centering
	\caption{\textbf{Average absolute prediction error of log$_2$ HI titer for test data across models and datasets.}}
	\label{errortable}	
	\makebox[\textwidth][c]{
	\small
	\begin{tabular}{ c c c c c c c c c c } 
	\hline
	\multicolumn{6}{c}{} 	&	\multicolumn{4}{c}{Test error}  \\
	Model	&	Data		&	Dimen 	& Location prior	& 	Serum effects 	&	Virus effects	& 	A/H3N2	& 	A/H1N1	& 	B/Vic	& 	B/Yam	\\
	\hline	
	1		&	HI			&	1D 		& Uninformed		&	Fixed 			&	None			&	1.35	&	0.94 	&	0.90	&	1.08	\\
	2		&	HI			&	2D 		& Uninformed		&	Fixed 			&	None			&	0.91 	&	0.78 	&	0.82	&	0.90	\\
	3		&	HI			&	3D 		& Uninformed		&	Fixed 			&	None			&	0.93 	&	0.80 	&	0.85	&	0.92	\\
	4		&	HI			&	4D 		& Uninformed		&	Fixed 			&	None			&	0.98 	&	0.84 	&	0.90	&	0.97	\\	
	5		&	HI			&	5D 		& Uninformed		&	Fixed 			&	None			&	1.04 	&	0.89 	&	0.98	&	1.04	\\	
	6		&	HI/year		&	2D 		& Drift				&	Fixed 			&	None			&	0.91	&	0.75	&	0.77	&	0.83	\\
	7		&	HI/year/seq	&	2D 		& Diffusion/Drift	&	Fixed 			&	None			&	0.89	&	0.74	&	0.74	&	0.83	\\
	8		&	HI/year/seq	&	2D 		& Diffusion/Drift	&	Estimated 		&	None			&	0.77	&	0.73	&	0.66	&	0.75	\\	
	9		&	HI/year/seq	&	2D 		& Diffusion/Drift	&	Fixed 			&	Estimated		&	0.80	&	0.72	&	0.69	&	0.75	\\		
	10		&	HI/year/seq	&	2D 		& Diffusion/Drift	&	Estimated		&	Estimated		&	0.76	&	0.71 	&	0.64	&	0.72	\\
	\hline
	\end{tabular}	
	}
\end{table}

@@

We further extend the model by estimating the strength of overall reactivity of each serum rather than fixing this at its maximum titer, and additionally, by estimating the strength of reactivity of each virus isolate.
We refer to these estimates as serum effects and virus effects, respectively.
These effects determine the baseline expectation of titer when virus and serum have identical antigenic locations.
We find that including these effects decreases test error further, resulting in comparable improvements from estimating virus and serum effects individually and giving an improvement of between 0.03 to 0.13 log$_2$ HI titers when estimating both virus and serum effects together (models 8--10; Table \ref{errortable}).
